---
title: "SURV625, HW-2"
author: "Sagnik Chakravarty"
format: pdf
header-includes:
  - \usepackage[a4paper, margin=0.5in]{geometry}
toc: FALSE
lot: FALSE
editor: visual
---

\newpage

\tableofcontents

\newpage

# Question 1

A stratified random sample of graduates from the academic departments at a small university is to be selected to estimate the mean starting salary of graduates from that university. The university has five academic departments, and preliminary data are obtained from an attempted census in a recent year:

| Stratum | Department       | $N_h$ | $\bar{Y}_h$ | $S_h$   |
|---------|------------------|-------|-------------|---------|
| 1       | Humanities       | 20    | 46,800      | 81,200  |
| 2       | Social Sciences  | 90    | 61,500      | 101,700 |
| 3       | Natural Sciences | 120   | 76,100      | 130,900 |
| 4       | Engineering      | 140   | 89,500      | 170,100 |
| 5       | Business         | 200   | 95,500      | 216,400 |

## a. What is the mean starting salary $\bar{Y}$ per graduate across all departments in the population, based on the data collected from this previous attempted census?

### Solution

```{r message=FALSE, warning=FALSE}
library(knitr)
library(dplyr)
library(kableExtra)
df <- data.frame(stratum = 1:5
                 ,Department = c('Humanities', 'Social Sciences', 'Natural Sciences', 'Engineering', 'Business'),
                 N_h = c(20, 90, 120, 140, 200),
                 Y_bar_h = c(46800, 61500, 76100, 89500, 95500),
                 S_h = c(81200, 101700, 130900, 170100, 216400))
df['W_h'] <- df$N_h/sum(df$N_h)
df['W_h.Y_h'] = df$W_h*df$Y_bar_h
kable(df, format = 'latex')
cat('the mean starting salary accross all departments:\t', sum(df$W_h.Y_h))
```

### Calculation

The mean starting salary per graduate across all departments in the population is the weighted mean of the stratum means, given by:

$$
\bar{Y} = \sum W_h\bar{Y}_h
$$

where:

-   $W_h = \frac{N_h}{N}$is the proportion of graduates in stratum $h$

-   $N_h$ is the number of graduates in department $h$

-   $\bar{Y}_h$ is the mean starting salary for department $h$

-   $N=\sum_h N_h$ is the total number of graduates.

$$
\begin{aligned}
&N = 20+90+120+140+200=570\\
&W_h = \frac{20}{570}, \frac{90}{570},\frac{120}{570},\frac{140}{570},\frac{200}{570}\\
&\bar{Y} = \frac{20}{570}\times46,800 + \frac{90}{570}\times61,500+\frac{120}{570}\times76,100+ \frac{140}{570}\times89,500 +\frac{200}{570}\times95,500 = 82864.91
\end{aligned}
$$

Hence $\bar{Y} = 82,864.91$

## b. What is the average within-stratum element variance $S_w^2 = \sum W_hS_h^2$

### Solution

```{r}
df['S_h^2'] = df$S_h^2
cat('The average within stratum element variance:\t', 
    sum(df$`S_h^2`*df$W_h))
```

### Calculation

$$
\begin{aligned}
S_h^2&= \frac{20}{570}\times81,200^2 + \frac{90}{570}\times101,700^2+\frac{120}{570}\times130,900^2+ \frac{140}{570}\times170,100^2 +\frac{200}{570}\times216,400^2\\
&=29009578070
\end{aligned}
$$

hence $2.9\times10^{10}$

## c. For a sample of $n = 100$, what is the proportionate allocation?

### Solution

```{r}
df['proportionate'] <- round(df$W_h *100)
kable(df[c('stratum', 'Department', 'proportionate')], format = 'latex')
```

### Calculation

In proportionate allocation, the sample size for each stratum $n_h$ is determined based on the proportion of the population in that stratum:

$$
n_h = W_hn
$$

where

-   $W_h = \frac{N_h}{N}$

-   n is the number of sample according to the question $n = 100$

$$
\begin{aligned}
&n_1 = \frac{20}{570}\times100 = 3.51\approx 4\\
&n_2 = \frac{90}{570}\times100 = 15.79\approx 16\\
&n_3 = \frac{120}{570}\times100 = 21.05\approx 21\\
&n_4 = \frac{140}{570}\times100 = 24.56\approx 25\\
&n_5 = \frac{200}{570}\times100 = 35.09\approx 35\\
\end{aligned}
$$

## d. For a sample of $n = 100$, what is the Neyman allocation?

### Solution

```{r}
df['W_hS_h'] = df$W_h*df$S_h
df['Neyman'] = round(100 * df$W_hS_h/sum(df$W_hS_h))
kable(df[c('stratum', 'Department', 'Neyman')], format = 'latex')
```

### Calculation

$$
n_h = k.W_hS_h = \frac{W_hS_h}{\sum_h W_hS_h}\times n \text{   where }k=\frac{n}{\sum_h W_hS_h}
$$

| Stratum | Department       | $W_hS_h$                                    | Calculation                             | $n_h$ |
|------------|------------|--------------------|------------------|------------|
| 1       | Humanities       | $\frac{20}{570}\times 81,200 = 2849.123$    | $\frac{2849.123}{164173.7}\times 100$   | 2     |
| 2       | Social Sciences  | $\frac{20}{570}\times 101,700 = 16,057.895$ | $\frac{16,057.895}{164173.7}\times 100$ | 10    |
| 3       | Natural Sciences | $\frac{20}{570}\times 130,900 = 27,557.895$ | $\frac{27,557.895}{164173.7}\times 100$ | 17    |
| 4       | Engineering      | $\frac{20}{570}\times 170,100 = 41,778.947$ | $\frac{41,778.947}{164173.7}\times 100$ | 25    |
| 5       | Business         | $\frac{20}{570}\times 216,400 = 75,929.825$ | $\frac{75,929.825}{164173.7}\times 100$ | 46    |

## e. Estimate the sampling variance of the mean for the proportionate c) and Neyman d) allocations.

### Solution

```{r}
kable(df %>%
  mutate('prop_variance' = W_h^2*(1-proportionate/N_h)*`S_h^2`/proportionate,
         'neyman_variance' = W_h^2*(1-Neyman/N_h)*`S_h^2`/Neyman) %>%
  summarize('Proportionate Variance' = sum(prop_variance),
            'Neyman Variance' = sum(neyman_variance)),
  format = 'latex', booktabs = TRUE) %>%
  kable_styling(latex_options = "scale_down")
```

### Calculation

$$
var(\bar{y}_w) = \sum_h W_h^2\frac{1-f}{n_h}s_h^2\text{ where }f=\frac{n_h}{N_h}
$$

| Stratum | Department       | Proportionate Variance                                             | Neyman Variance                                                    |
|-------------|-------------|------------------------|------------------------|
| 1       | Humanities       | $(\frac{20}{570})^2 (1-\frac{4}{20})\frac{81200^2}{4} = 1,623,500$ | $(\frac{20}{570})^2 (1-\frac{2}{20})\frac{81200^2}{2} = 3,652,875$ |
| 2       | Social Sciences  | 13,250,932                                                         | 22,920,532                                                         |
| 3       | Natural Sciences | 29,835,047                                                         | 38,344,151                                                         |
| 4       | Engineering      | 57,351,500                                                         | 57,351,500                                                         |
| 5       | Business         | 135,897,259                                                        | 96,506,749                                                         |

hence

-   **Proportionate Variance** = 237958239

-   **Neyman Variance** = 218775808

## f. Estimate the total element variance $S^2$

### Solution

```{r}
df %>%
  mutate('S2' = (N_h-1)/(sum(N_h) - 1)*(`S_h^2` + (Y_bar_h - 82864.91)^2)) %>%
  summarise('S^2' = sum(S2)) %>% kable(format = 'latex')
```

### Calculation

$$
S^2 = \sum_h \frac{N_h-1}{N-1}S_h^2+\sum_h \frac{N_h-1}{N-1}(\bar{Y}_h-\bar{Y})^2 = 29058520065
$$

hence $S^2 = 2.9\times10^{10}$

## g. What are the design effects of the proportionate and Neyman allocations?

### Solution

```{r}
cat('Design Effect of Proportion Allocation:\t', 237958239/2905852006,
    '\nDesign Effect of Neyman Allocation:\t', 218775808/2905852006)
```

### Calculation

$$
deff = \frac{var(\bar{y})}{var_{SRS}(\bar{y})} = \frac{s_w^2}{s^2}
$$

hence

-   Design Effect of Proportion Allocation: **0.08188932**

-   Design Effect of Neyman Allocation: **0.07528801**

## h. Suppose that the cost-per-element was not the same in each stratum: $C_1 = C_2= C_3=\$30, C_4=C_5=\$40$

### 1. The client requesting a stratified sample design has indicated that the total available data collection budget is 𝐶 = \$5,000, with the stratum specific costs per element listed above. What allocation will minimize the sampling variance of the mean under these cost constraints?

#### Solution

```{r}
df['Cost'] <- c(30, 30, 30, 40, 40)
k_cost <- 5000/sum((df$W_h*df$S_h*sqrt(df$Cost)))
df <- df %>% 
  mutate('Cost Allocation' = round(k_cost*W_h*S_h/sqrt(Cost))) 
df %>%
  select(stratum, Department, Cost, `Cost Allocation`) %>%
  kable(format = 'latex')
```

#### Calculation

$$
\begin{aligned}
&C = 5000\\
&k = \frac{C}{\sum_hW_hS_h\sqrt{C_h}} =\frac{5000}{0.035\times81200\times\sqrt{30}+0.157\times101700\times\sqrt{30}+\cdots} = 0.005005233\\
&n_h = \frac{kW_hS_h}{\sqrt{C_h}} = (\frac{0.005\times 0.035\times81200}{\sqrt{0.005}}, \cdots) =(3, 15, 25, 33, 60)
\end{aligned}
$$

### 2. Estimate the expected sampling variance and design effect of the mean starting salary under this allocation

#### Solution

```{r}
df %>%
  mutate('cost_variance' = W_h^2*(1-`Cost Allocation`/N_h)*`S_h^2`/`Cost Allocation`) %>%
  summarize('Cost Allocation Variance' = sum(cost_variance),
            'design effect' = sum(cost_variance)/2905852006) %>%
  kable(format = 'latex')
```

#### Calculation

$$
var(\bar{y}_w) = \sum_h W_h^2\frac{1-f}{n_h}s_h^2 = 148362056
$$

$$
\text{Design Effect} =  \frac{var(\bar{y})}{var_{SRS}(\bar{y})} = \frac{s_w^2}{s^2} = \frac{148362056}{2905852006} = 0.051
$$
